Type Error Telemetry at Scale: Pseudonymized and Stochastic

### 2023-09-21 programming23/paper.tex
stochastic
sk edits: less marketing, clarify Luau vs types vs nocheck distinction

\title{Type Error Telemetry at Scale: Pseudonymized and Stochiastic}
% lean, lite, modest, non-intrusive, private, ...
% {Millions of Type Errors} = 72M forced strict, 1/2M type error is disappointing!
%% \subtitle{Impersonal Telemetry to Measure User Experience}

@@

  \paragraph{Context}
  %% What is the broad context of the work? What is
  %% the importance of the general research area?
  {Roblox Studio} lets millions of creators
  build interactive experiences by programming in a variant
  of Lua called Luau.
  The creators form a broad group, ranging from novices writing
  their first script to professional developers, thus Luau
  must support a wide audience.
  As part of its efforts to support all kinds programmers, Luau includes a
  gradual type system and goes to great lengths to minimize false positive
  errors.

  \paragraph{Inquiry}
  %% What problem or question does the paper
  %% address? How has this problem or question been
  %% addressed by others (if at all)?
  Since Luau is in the hands of many creators, we want to collect their feedback
  and ultimately improve the language design (in particular, the type system).
  We want feedback at a large scale, as creators work on their actual projects,
  which suggests the deployment of client-side telemetry.
  %% ^ awkward
  However, we must respect creators privacy.
  Whatever telemetry we use to study the Luau language cannot include
  personally-identifiable information.
  For example, we cannot monitor source code or error messages as they
  may contain personal data.
  Finding ways to harness telemetry in a private way is the central research challenge.

  \paragraph{Approach}
  %% What was done that unveiled new knowledge?
  We designed and implemented a privacy-preserving telemetry system for Luau.
  Telemetry records include a timestamp, a (randomized) session id, a reason
  for sending, and a sequence of counter values summarizing the results of type
  analysis.
  Over four months in Spring 2023, we deployed the telemetry system and collected
  1.5 million telemetry records.
  We analyzed the data to improve Luau.

  \paragraph{Knowledge}
  %% What new facts were uncovered? If the
  %% research was not results oriented, what new
  %% capabilities are enabled by the work?
  We present several findings about Luau:
  some confirm hypotheses by the Luau team, others are rather surprising, but
  all findings support our claim that privacy-preserving telemetry is a useful
  way to collect feedback.
  One expected finding is that types are unpopular; there is an order-of-maginitude
  difference in the number of sessions that use different type analysis modes.
  One surprise is that the strict type analysis mode evidently makes it difficult
  to connect a script with data assets.
  Lastly, one reassuring finding is that type analysis rarely hits its internal
  limits on problem size.

@@

{Roblox} is a platform for {shared virtual experiences}, with
65~million Daily Active Users, and 14~billion hours of engagement in
April--June 2023; there are 3~million creators using {Roblox Studio},
creating 5~million experiences~\cite{corp.roblox.com}.

{Roblox experiences} are scripted using the 
{Luau} programming language~\cite{luau-lang.org},
an extension of {Lua~5.1~\cite{lua}}.
The main extension is the addition of a static type system, which uses
type inference to synthesize types for user code on the fly.
These types are used primarily in type-driven tooling such as autocomplete and
API documentation~\cite{luau-autocomplete}, but creators can also opt in to
receiving type error reports.

As discussed in~\cite{bfj-hatra-2021},
the goals of the {Luau} type system are rather different from
a traditional type system, which focuses on compilation and memory safety.
{Luau} has a diverse user community, ranging from
students in code camps to professional development studios. These
creators have quite different needs, with different emphases on
enabling rapid creation and ensuring software quality.
{Luau} therefore supports three typing modes---\mstrict{} types,
relaxed (\mnonstrict{}) types, and no types (\mnocheck{})---and lets users switch
between modes gradually~\cite{st-sfp-2006,tfffgksst-snapl-2017}, one module at a time.

@@

Most users of {Roblox Studio} do not opt in to type error
reporting, and so they do not see the ``squiggly underlining'' that
indicates a type error site. Nonetheless, the type inference system
still runs in the background (since it drives autocomplete and other
type-based tools), letting us see which type errors would have
been reported and whether users fix these errors over time.

@@

Analysis pipeline to be freely available.
Data may be available, depends on Roblox.

@@

FILL table = object

\begin{figure}

  \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
  \includegraphics[width=.45\textwidth]{img/roblox-studio-ide.png}

@@

Creators of {Roblox experiences} use {Roblox Studio},
which combines {3D creation} tools as well as an Integrated
Developer Environment (IDE), as seen in Fig.~\ref{fig:roblox-studio}.
The IDE includes an optional ``Script Analysis'' widget, which
reports syntax errors, type errors, and problems identified by
lint tools. The script editor also (optionally) highlights
the location in code where reported errors occur.
To opt in to type error reporting, creators set a \emph{mode}
for each script, which is one of:
\begin{itemize}
  \item \mnocheck{}: only syntax errors are reported,
  \item \mnonstrict{}: all syntax errors, and a subset of ``high confidence'' type errors, are reported, or
  \item \mstrict{}: all syntax errors and type errors, are reported.
\end{itemize}
As an example of nonstrict mode, the following program only reports one error:
\begin{verbatim}
  --!nonstrict
  local x = { p = 5, q = nil }
  if condition then x.q = 7 end
  local y = x.p + x.q --> no type error
  local z = x.r       --> "Key 'r' not found in table 'x'"
\end{verbatim}
but in strict mode it reports two:
\begin{verbatim}
  --!strict
  local x = { p = 5, q = nil }
  if condition then x.q = 7 end
  local y = x.p + x.q --> "Type 'nil' could not be converted into 'number'"
  local z = x.r       --> "Key 'r' not found in table 'x'"
\end{verbatim}
In cases like this, where it is undecidable whether there will be a run-time error,
strict mode errs on the side of reporting an error, and nonstrict mode errs on
the side of suppressing the error.

@@

The state of the world in a {Roblox} experience is captured by
the \emph{data model}, which is a tree of \emph{instances}, such as
parts, models, meshes, effects, lighting, audio assets, and physics
constraints such as forces, springs and joints.
The data model can be seen in the Explorer view of~\Cref{fig:roblox-studio}.
While an experience is under development, it is typical for the data
model to be edited (for example instances to be added, deleted, moved
or renamed). Since the initial shape of the data model tree is reflected in
the type system, it is possible for these edits to introduce type errors.

@@

  %% FILL examples for each error?
  \begin{tabular}{ll}
    Label & Interpretation \\\midrule
    \code{CodeTooComplex} & Type analysis failed, cannot understand the code\!\!\! \\
    \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit\!\!\! \\
    \code{SyntaxError} & Basic parse error, e.g., \code{for if end} \\

@@

In order to avoid swamping our telemetry servers, we randomly throttle
the generation of telemetry records:
\begin{itemize}
  \item
    1\% of Roblox Studio sessions were enrolled in generating telemetry, and
  \item
    0.5\% of runs of the typechecker (approximately 1 in 200 keystrokes)
      in an enrolled session generated a telemetry record.
\end{itemize}
In order to find how file-switching correlated with type errors, we additionally have:
\begin{itemize}
  \item
    every run of the typechecker in an enrolled session which switched file generated a telemetry record.
\end{itemize}
Each telemetry record contains:
\begin{description}
  \item[D1] metadata: a psedonymized session identifier, timestamp and session duration,
  \item[D2] mode of the current file: (\mnocheck{}, \mnonstrict{}, or \mstrict{}),
  \item[D3] reason for sending: (random selection, or file switch),
  \item[D4] size of codebase: number of files, and number of lines of code type checked,
  \item[D5] size of edit range
  \item[D6] overall type errors: for previous and current states, we have the total number, the nunber in the current file, and number in the edit range,
  \item[D7] specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
    \FILL{} explain previous/current in detail, explain that this data takes up most of the telemetry record
  \item[D8] forced strict errors: specific errors for the
    edit range computed as if the user had strict mode enabled.
  \item[D9] TooComplexErrors: project-wide count for this one specific error type
\end{description}
To track the edit range, we record a start and end position, which we
update appropriately on every edit. This can result in very large edit
ranges, for example, if the user edits at the beginning and end of the
file.
The main benefit of this strategy is that it ensures constant-size telemetry records.

@@

\begin{figure}[t]
  %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days

>>>>

\title{Type Error Telemetry at Scale: Pseudonymized and Stochastic}
% lean, lite, modest, non-intrusive, private, ...
% {Millions of Type Errors} = 72M forced strict, 1/2M type error is disappointing!
%% \subtitle{Impersonal Telemetry to Measure User Experience}

@@

  \paragraph{Context}
  %% What is the broad context of the work? What is
  %% the importance of the general research area?
  {Roblox Studio} lets millions of creators
  build interactive experiences by programming in a variant
  of Lua called Luau.
  The creators form a broad group, ranging from novices writing
  their first script to professional developers, thus Luau
  must support a wide audience.
  As part of its efforts to support all kinds programmers, Luau includes an
  optional, gradual type system and goes to great lengths to minimize false
  positive errors.

  \paragraph{Inquiry}
  %% What problem or question does the paper
  %% address? How has this problem or question been
  %% addressed by others (if at all)?
  Since Luau is currently being used by many creators, we want to collect their feedback
  to improve the language and, in particular, the type system.
  The standard way to collect feedback at a large scale from creators working on
  realistic projects is to deploy client-side telemetry; however, we must do
  so in a way that respects creators' privacy.
  Whatever telemetry we use to study the Luau language cannot include
  personally-identifiable information.
  For example, we cannot monitor source code or error messages as they
  may contain personal data.
  Finding ways to harness telemetry in a private way is the central research challenge.
  %% TODO awkward

  \paragraph{Approach}
  %% What was done that unveiled new knowledge?
  We designed and implemented a pseudonymized telemetry system for Luau.
  Telemetry records include a timestamp, a session id, a reason for sending,
  and a numeric summary of the most recent type analysis.
  This information lets us follow type analysis over time within a codebase
  without revealing private information.
  Over four months in Spring 2023, we deployed the telemetry system and collected
  1.5 million telemetry records.

  \paragraph{Knowledge}
  %% What new facts were uncovered? If the
  %% research was not results oriented, what new
  %% capabilities are enabled by the work?
  We present several findings about Luau.
  Some confirm hypotheses by the Luau team while others are rather surprising,
  but all findings support our claim that pseudonymized telemetry is a
  useful way to collect feedback.
  One of the less-surprising findings is that types are unpopular:
  there is an 100x gap between the number of untyped Luau sessions and the number
  of typed ones.
  One surprise is that the strict mode for type analysis is very conservative
  about interactions with data assets.
  Instead of using the flexible dynamic type, it uses a top type and requires
  creators to down-cast assets.
  Lastly, a reassuring finding is that type analysis rarely hits its internal
  limits on problem size.

@@

{Roblox} is a platform for {shared virtual experiences}, with
65~million Daily Active Users, and 14~billion hours of engagement in
April--June 2023.
There are over 5~million experiences available on the platform thanks to a
worldwide community of 3~million creators~\cite{corp.roblox.com}.

Creators build experiences using the
{Luau} programming language~\cite{luau-lang.org},
an extension of {Lua~5.1~\cite{lua}}.
The main addition in Luau is a static type system, which infers
types for all Luau programs on the fly, as creators modify the code.
These types are used primarily in type-driven tooling such as autocomplete and
API documentation~\cite{luau-autocomplete}; the typical creator is unaware
that a type system powers these tools.
However, creators may write their own types, and they can opt in
to receiving ype error reports.

As discussed in~\cite{bfj-hatra-2021},
the goals of the {Luau} type system are rather different from
a traditional type system, which focuses on compilation and memory safety.
{Luau} has a diverse user community, ranging from
students in code camps to professional development studios. These
creators have quite different needs, with different emphases on
enabling rapid creation and ensuring software quality.
{Luau} therefore supports three typing modes for user code:
the hands-off default \mnocheck{},
a relaxed \mnonstrict{} mode,
and a rigorous \mstrict{} mode.
Creators can switch between modes
gradually~\cite{st-sfp-2006,tfffgksst-snapl-2017}, one module at a time.

@@

Most users of {Roblox Studio} do not opt in to type error
reporting (\pct{90},~\cref{s:data}), and so they do not see the ``squiggly underlining'' that
indicates a type error site. Nonetheless, the type inference system
still runs in the background (since it drives autocomplete and other
type-based tools), letting us see which type errors would have
been reported and whether users fix these errors over time.

@@

The data and the analysis pipeline are freely available.
\FILL{} link.

@@

\begin{figure}[t]\centering
  \includegraphics[width=.45\textwidth]{img/roblox-studio.png}
  \includegraphics[width=.45\textwidth]{img/roblox-studio-ide.png}

@@

%% \FILL{} explain the name "creators"?

{Roblox Studio} is a workbench that
combines {3D creation} tools and an Integrated
Developer Environment (IDE), as seen in \Cref{fig:roblox-studio}.
The IDE includes an optional ``Script Analysis'' widget, which
reports syntax errors, type errors, and problems identified by
lint tools. The script editor also (optionally) highlights
the location in code where reported errors occur.
To opt in to type error reporting, creators set a \emph{mode}
for each script, which is one of:
\begin{itemize}
  \item \mnocheck{}: only syntax errors are reported,
  \item \mnonstrict{}: all syntax errors, and a subset of ``high confidence'' type errors, are reported, or
  \item \mstrict{}: all syntax errors and type errors, are reported.
\end{itemize}
As an example of nonstrict mode, the following program only reports one error:
\begin{verbatim}
  --!nonstrict
  local x = { p = 5, q = nil }
  if condition then x.q = 7 end
  local y = x.p + x.q --> no type error
  local z = x.r       --> "Key 'r' not found in table 'x'"
\end{verbatim}
but in strict mode it reports two:
\begin{verbatim}
  --!strict
  local x = { p = 5, q = nil }
  if condition then x.q = 7 end
  local y = x.p + x.q --> "Type 'nil' could not be converted into 'number'"
  local z = x.r       --> "Key 'r' not found in table 'x'"
\end{verbatim}
In cases like this, where it is undecidable whether there will be a run-time error,
strict mode errs on the side of reporting an error, and nonstrict mode errs on
the side of suppressing the error.

@@

The state of the world in a {Roblox} experience is captured by
the \emph{data model}, which is a tree of \emph{instances}, such as
parts, models, meshes, effects, lighting, audio assets, and physics
constraints such as forces, springs and joints.
%% The data model can be seen in the Explorer view of~\Cref{fig:roblox-studio}.
%% ben: where is it? very hard to see
While an experience is under development, it is typical for the data
model to be edited (for example instances to be added, deleted, moved
or renamed). Since the initial shape of the data model tree is reflected in
the type system, it is possible for these edits to introduce type errors.

%% FILL table = object

@@

  %% FILL examples for each error?
  \begin{tabular}{ll}
    Label & Interpretation \\\midrule
    \code{CodeTooComplex} & Type analysis failed, cannot understand the code\!\!\! \\
    %% \code{UnificationTooComplex} & Type analysis failed, unification solver hit a limit\!\!\! \\
    \code{SyntaxError} & Basic parse error, e.g., \code{for if end} \\

@@

In order to avoid swamping our telemetry servers, we randomly throttle
the generation of telemetry records:
\begin{itemize}
  \item
    1\% of Roblox Studio sessions were enrolled in generating telemetry, and
  \item
    0.5\% of runs of the typechecker (approximately 1 in 200 keystrokes)
      in an enrolled session generated a telemetry record.
\end{itemize}
In order to find how file-switching correlated with type errors, we additionally have:
\begin{itemize}
  \item
    every run of the typechecker in an enrolled session which switched file generated a telemetry record.
\end{itemize}
Each telemetry record contains nine kinds of data:
record 
\begin{enumerate}
  \item metadata: a psedonymized session identifier, timestamp and session duration,
  \item mode of the current file: (\mnocheck{}, \mnonstrict{}, or \mstrict{}),
  \item reason for sending: (random selection, or file switch),
  \item size of codebase: number of files, and number of lines of code type checked,
  \item size of edit range
  \item overall type errors: for previous and current states, we have the total number, the nunber in the current file, and number in the edit range,
  \item specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
    \FILL{} explain previous/current in detail, explain that this data takes up most of the telemetry record
  \item forced strict errors: specific errors for the
    edit range computed as if the user had strict mode enabled.
  \item TooComplexErrors: project-wide count for this one specific error type
\end{enumerate}
To track the edit range, we record a start and end position, which we
update appropriately on every edit. This can result in very large edit
ranges, for example, if the user edits at the beginning and end of the
file.
The main benefit of this strategy is that it ensures constant-size telemetry records.

@@

\begin{figure}[t]\centering
  %% NOTE max time = 1 million seconds = 15 days ...  +100 are over 2 days
